Composition-climate paper

### 2016-11-25 Response_to_reviewers.Rmd

In the results we note:

*Analysis included in the supplemental material indicates that a shift to an earlier baseline (using the estimated change rate from [@burnette2010daily] in Manhattan, Kansas) results in no change in this pattern (see "R/confounding_table.R" in S1).*

> Another issue is lack of discussion about the time period when the forests sampled by FIA got established. It is typical in SDM studies to test the strength of models from different time periods in tree SDM work and select the time period that is most strongly associated with species presence.  In the western US, we found 1950-1980 to produce stronger models than 1980-2010.

We note that this is not an excersise in creating SDMs, a point we will return to later, but take the point of the association between vegetation and climate varies between establishment and long term survival of forest stocks, we believe that our contribution further underscores this point. The normals for the region, beginning with a normal ending 1980 and continuing to a normal ending 2014, give a total range of 0.35^o^C for $T_{max}$, 0.84^o^C for $T_{min}$, 1.15^o^C for $T_{diff}$ and 20mm for $P_{ann}$.  These estimates are all lower that the variance within any one of the normals (1.5^o^C, 3.0^o^C, 3.1^o^C and 78mm) respectively, but, as pointed out in Baker *et al*. (2016; doi:10.1111/gcb.13273) uncertainty in the underlying estimates are often overlooked.  Given the variability within windows, and the relatively young age of most stands in WI, MN and MI (FIA data shows mean ages of 52, 56 and 49 years respectively), we believe that the choice of normal is acceptable.  Having said that, we examined the importance using both shifts in PLS era, and shifts in the modern climate normal.  The largest effect is in $t_{max}$, since the most recent normals all show a significant increase in $t_{max}$.

![](figures/corrected_all.png)

@@

> A third issue is the role of climate variability over the 1895 to 2014 time period.  Decadal oscillations are common in this period in many parts of the US.  Trees are likely integrating over 2 or more of these climate cycles.  SDMs should be done to best represent the climate and variability that selected for tree presence when the trees were measured.  Your analysis would be strengthened with statistical analysis of climate patters from 1895 to present for your predictor variables and an analysis of which time period produces the best SDMs.

I believe that this is addressed in the response to the prior comment.

@@

> 195 It is important to discuss why univariate climate tree presence functions are relevant to the climate factors that strongly influence tree species presence or absence.  SDM modeling typically shows that a particular combination of climate and habitat factors is much more strongly associate with tree presence and any random climate variable.  Your analysis would be stronger to do model selection for best multivariate SDM functions and then compare these functions between historic and present.

We apreciate this point and thank the reviewer for bringing it up.  The challenge in this case is that a multivariate SDM blends the climate variables. Further, a multi-model SDM approach might weight climate variables differently across models, through time, making sensitivity analysis for the impact of each climate variable challenging.  We point out that this analysis is not intended to focus on the construction of SDMs, but rather to show the relationship between land use and the individual climate variables, through the distribution of tree species.  Thus, disentagling land use and climate shifts on a set of multivariate combination of variables from each time slice becomes extremely challenging, and, is effectively a separate paper.

What this paper intends to point to is that land use is impacting climate variables differentially across species as a result of land use, and that this may have implications for SDM construction.  Shifts in the strength of SDMs between time scales and potential utility of climate variables in SDM construction is a separate analysis and paper.

@@

> 211-218 The analyses should control for spatial autocorrelation, simple t-tests would seem to have an inflated sample size due not including spatial autocorrelation in the model.

We have included spatial autocorrelation in the t-tests using the `SpatialPack` function`modified.ttest`.  We use confidence intervals based on the raw t.test scores, but report significance based on the modified.ttest calculation.

### 2016-11-27 Composition_Ecology.Rmd
Adding the changes related to normals & SDMs from reviewer 2.
Began removing some of the *pathogen* content, or at least it's emphasis in the purpose of the paper.
Began changing tense in the methods to present where appropriate based on the advice of Reviewer 3.

Better explanation of the terminology later used in figure captions, based on the advice of Reviewer 3.

Explained why we used the normal we did.

Added in the Hellinger equation and rephrased the explanation of how it was used.

Added some packages into the **Analysis** section.

Added references to Rhemtulla et al 2007 in the Discu

Contemporary forest inventory data are widely used to understand environmental controls on tree species distributions and to project forest responses to climate change, but the stability and representativeness of contemporary tree-climate distributions is poorly understood. Here we show that tree-climate relationships for 15 common tree genera in the north-central US have significantly altered over the last two centuries due to historical land use, pathogen outbreaks, and climate change. Realized niches have shifted towards higher minimum temperatures, cooler maximum temperatures, and higher rainfall. A new attribution method implicates both historical climate change and land use in these shifts, with the relative importance varying among genera and climate variables. Most climate/land-use interactions are compounding, in which historical land-use reinforces climatic shifts in species distributions toward cooler or wetter distributions. The prevalence of compounding interactions imply that contemporary-based models of species distributions may underestimate species resilience to climate change.

@@

Tree distributions and community composition in the upper Midwestern US have been heavily influenced by land use conversion associated with EuroAmerican Settlement in the 19^th^ and 20^th^ Centuries [@goring2016preset;@schulte2007homogenization;@hanberry2012comparison;@rhemtulla2009historical], outbreaks of exotic pathogens [@barnes1979stem], climate changes accompanying the end of the Little Ice Age [@hotchkiss2007response] and current anthropogenic warming [@wicci2011climate]. Modern forests, as represented by FIA surveys [@woudenberg2010forest], show greater homogeneity, weaker ecotones, and significantly different species associations than forests surveyed as part of the Public Land Survey System (PLS) prior to widespread land-use conversion in the late 1800s [@goring2016preset;@schulte2007homogenization]. Wildland ecosystems have been converted to agricultural and urban land use [@rhemtulla2009historical], particularly in historical prairie and savanna ecosystems (`r fig_nums("land_use_clim", display = "cite")`f).  In northern forested regions, ranges of early successional species have expanded and abundances of late-successional species are reduced due to widespread logging in the 19^th^ and 20^th^ century, and ongoing land management [@schulte2007homogenization;@goring2016preset].  Additionally, the formerly dominant *Ulmus americana* has experienced rangewide population collapses caused by the introduction of exotic pathogens [@barnes1979stem], as did *Castanea dentata*, with other tree species currently threatened (*Tsuga canadensis*, *Fraxinus* spp.)

Here we demonstrate significant shifts in the climatic distributions of tree genera in the upper Midwest over the last 100 years as a result of historical climate change, land use, and pathogen outbreaks, and we diagnose the relative influence of historical climate and land use changes on these shifts.  We use gridded historical and contemporary climate observations (PRISM-LT) and estimates of pre-settlement and modern forest vegetation [@goring2016preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwestern United States.  We construct four observed and hypothetical sets of tree-climate relationships, using a 2x2 factorial of pre-settlement and modern vegetation and climates.  We develop a new method, based on Hellinger distances among tree-climate distributions, to quantify the total change and attribute the relative importance of historical vegetation and climate change. We assess whether climate and vegetation changes are compounding, counteracting or confounding based on the magnitude and direction of species-climate shifts. We use this framework to discuss the potential impacts of historical land-use and climate change on inferences about species resilience to climate change, as predicted by species-distribution models based soley on modern distributional data.

@@

Estimates of settlement-era forest composition and tree distribution are based on survey data from the PLS [@white1983history;@schulte2001original;@goring2016preset], aggregated to a 64 km^2^grid across the upper Midwestern United States [@goring2016preset]. We briefly review the methods here. PLS data originally were collected on a 1.61 x 1.61 km (1 x 1 mi) grid across the Upper Midwest from 1830 to   1910 [@schulte2001original;@almendinger1996minnesota;@white1983history]. At each survey point surveyors noted the closest two to four trees and recorded their common name, distances from survey points, diameters at breast height, and azimuths to each tree. Goring et al. [-@goring2016preset] aggregated this data to an 8x8 km grid, resulting in a gridded data product with an average of 61 survey points per cell, or approximately 120 trees per cell. This process included data cleaning, taxonomic standardization, and the application of spatially varying correction factors [@goring2016preset]. Using this data, we produce presence-absence distributions for each taxon, across the upper Midwest for 15 common tree genera. A few common names, such as "Ironwood" can not be clearly resolved to a single tree genera and so we combine *Ostrya* and *Carpinus*, and *Thuja* and *Juniperus*. 'Presence' is defined as at least one recorded tree in any 8x8 km grid cell and 'range' is considered to be the extent in climate space of all cells with recorded presence for a specific taxon.

@@

Comparisons between historical and modern vegetation must address differences in sampling design between the PLS and FIA datasets [@goring2016preset;@kronenfeld2014validating;@kronenfeld2010influence]. Here, the primary solution for maximizing comparability is via aggregation, both spatially, by aggregating records from both datasets to a common scale (8x8 km), and taxonomically by aggregating taxa to the level of genera. Nonetheless, differences persist.  The FIA contains fewer plots per 8 km cell, with more trees per plot than the PLS data and FIA observations are collected only from forested or partially forested landscapes [@woudenberg2010forest], causing many tree taxa to have smaller spatial extents in the FIA than in the PLS datasets. Scattered trees may be present in regions without permanent or temporary FIA plots, complicating the ability to assign 'absence' within the FIA. However, the practice of using FIA data as a surrogate for tree distribution data is widespread [@iverson2013tree;@iverson1998predicting], so these complications are general. In prior comparative analyses of the PLS and FIA data, Goring et al. [-@goring2016preset] tested the effect of differential sampling design on the mapped patterns of novel and disappearing forests, and found little effect. FIA sampling represents a system with the potential for high heterogeneity within cells because of intensive sampling at the plot level, and, with one to few plots within an 8 x 8km grid cell, high between cell heterogeneity [@goring2016preset]. PLS sampling is likely to have low heterogeneity within and between cells because of the uniform spatial distribution, and the high number of points within cells [@goring2016preset].  Detectability at the plot level for FIA data is likely to be higher than for the PLS, however, given the variable radius of PLS plots, detectability of trees at forest margins is likely to be higher in the PLS, and PLS sample plots are likely to sample higher landscape heterogeneity within a given cell because of their uniform distribution across the landscape.

For all species, the range within the Upper Midwest represents an incomplete sample of the species range, however, the prairie-forest boundary along the western margin represents a southern and western range limit for many species, and represents a climatic space that is warmer and drier than elsewhere in the region.

@@

Historical and modern climate data are derived from monthly PRISM LT data for 1895 to 2014 CE [@prism2004]. The PRISM climate variables are available at 800m resolution and have been interpolated from station locations using elevation, aspect and other data.  Estimates for mean daily July temperature ($T_{max}$), mean daily January temperature ($T_{min}$), annual temperature range ($T_{diff}$: $T_{max}$ - $T_{min}$), and annually summed daily precipitation ($P_{ann}$) were resampled to the 8 x 8km grid used for PLS and FIA data. These climate variables were chosen because of their relative importance as demonstrated in prior research [@iverson1998predicting] and their use in past climate reconstruction [@maiorano2013building].  Recent statistical advances have opened the possibility of obtaining temperature and precipitation estimates from historical observations as early as the 1830s [@tiptoninreview], however the behaviour of uncertainties for derived variables such as potential evapo-transpiration, which ranked highly in previous work [@iverson1998predicting] are not currently known.

Two periods were selected for calculation of climate normals: 1895 to 1919 for the historical period and 1990 - 2014 for the modern period.  There is an unavoidable temporal mismatch between the historical climate data and the PLS survey period, on the order of 50 to 70 years; few meteorological records exist in this region prior to 1895 CE, and no gridded dataset exists at high spatio-temporal resolution for this time period for all climate variables.  The effects of temporal mismatch should be minor given that Northern Hemisphere temperature trends were generally small during the 19^th^ century [@mann2009global].  To check whether an 1895 to 1919 CE climate normal is a reasonable surrogate for pre-settlement climates we compare the instrumental record against dendroclimatic reconstructions of the Palmer Drought Severity Index [@heim2002review] (`r fig_nums("land_use_clim", display = "cite")`), and test July temperature reconstructions from observer data against our $T_{max}$ (Supplement 1).  Baker et al. [-@baker1985minnesota] show long term increases in mean annual temperature at Ft. Snelling and other spatially proximate records, while corrected estimates from Fort Leavenworth in Kansas [@burnette2010daily] show increasing temperatures in all seasons, with faster increases during winter (0.10^o^C decade^-1^) than during summer (0.07^o^C decade^-1^).  This implies that historical estimates may be biased by +0.35^o^C for $t_{max}$, by +0.5^o^C for $t_{min}$ and +0.15^o^C for $t_{diff}$.

@@

Distributions of tree genera within climate space are described as univariate probability density functions (PDFs) for the pre-settlement and modern eras.  Differences among PDFs are quantified using Hellinger distances [@matthes2016benchmarking] for the four combined states of vegetation and climate ($V_{H}C_{H}$, $V_{H}C_{M}$, $V_{M}C_{H}$, $V_{M}C_{M}$, where V represents vegetation, C represents climate, and M and H represent modern and historical conditions). The total shift ($d_{tot}$) is calculated as the Hellinger distance between $V_{M}C_{M}$ and $V_{H}C_{H}$ (`r fig_nums("fourPanel", display = "cite")`a: top left to bottom right). The effect of historical climate change on shifting climate distributions of tree genera is $d_{c}$, the difference between PLS and FIA tree distributions, overlaid on late 19th-century climates ($V_{H}C_{H}$ versus $V_{H}C_{M}$; `r fig_nums("fourPanel", display = "cite")`a; left column). The difference between the PLSS and modern tree distributions ($d_{v}$) when overlaid on historical climate ($V_{H}C_{H}$ versus $V_{M}C_{H}$) is attributed largely to land use change, as described above, given the widespread post-settlement disturbance in the upper Midwest; Figure 2a; top row).

@@

All analysis uses R [@Team2014], RMarkdown [@rmarkdown2016] and R packages `colorspace`, `captioner`, `raster`, `ggplot2`, `mgcv`, `reshape2`, `plyr`, `gridExtra`, `rgdal`, `RColorBrewer`, `analogue`, and `maptools`. All code is hosted on GitHub (http://github.com/PalEON-Project/Composition_Climate). Maps use the public domain Natural Earth Data (http://naturalearthdata.com) products for political and landscape features.

@@

`r fig_nums("boxplots", "Changes in the realized climate distributions for tree genera in the upper Midwestern United States. In the box-and-whisker plots, the central line indicates the median, box margins indicate 25^th^ and 75^th^ percentiles, whisker tips indicate 5^th^ and 95^th^ percentiles, and dots indicate outliers beyond the 95^th^ percentile. These shifts in realized climate distributions are produced by the joint effects of historical climate change, pathogen-induced tree mortality, and land-use change.")`

@@

`r fig_nums("hellPlot", "The total shift in climatic distribution over the last 100 to 150 years for each tree genus ($d_{tot}$) plotted against the amount of shift caused by historical climate change vs. changes in tree distributions ($d_{c} - d_{v}$), the latter mainly due to historical land-use and pathogen outbreaks.  Results are plotted separately for each of the four climate variables considered here. Shifts in distributions are calculated as the Hellinger distance between tree-climate distributions for different combinations of pre-settlement and modern vegetation and climate (Fig. 2).  For $d_{c}-d_{v}$ (Y-axis), negative values indicate a stronger influence of historical land-use on shifting climate distributions, while positive values indicate a stronger influence of historical climate change on shifting distributions.")`

@@

`r tab_nums("compTable", "Compounding ($\\Box$), confounding (-) and counteracting ($\\bigcirc$) relationships between land-use and climate change.  When the land-use/climate interaction was regressed against spatial vaariables (latitude, longitude) no significant relationship was found.")`

@@

pander::pandoc.table(conf_table, justify = "left", style = "rmarkdown")
knitr::kable(conf_table_corr)
```

@@

In this study, shifts in realized climate niches and the underlying interactions between climate change and land use emerge from three basic processes.  First, climates have changed over the 20^th^ and early 21^st^ centuries, with rising winter minimum temperatures, declining summer temperatures, and higher precipitation.  These climate changes are produced by a combination of natural variability and anthropogenic change; the latter signal is strongest in the second half of the 20^th^ century and early 21^st^ century [@estrada2013statistically]. Second, tree species distributions have not yet fully adjusted to these climate changes, causing tree distributions to be in partial disequilibrium with climate [@svenning2013disequilibrium;@williamsinpress]. Third, anthropogenic land use has selectively eliminated portions of geographic and environmental space from species' potential ranges. Agricultural conversion has largely eliminated open forests in warm and dry regions, while summer temperatures have declined and precipitation increased, resulting in a compounding effect; species occupy cooler habitats with higher precipitation than during historical times. Intensification of land-use in the northern prairie-forest margin shifts species further than expected along the precipitation gradient. While fire and logging have played a role in changing distribution patterns in the north and northeast, they have not resulted in the exclusion or extirpation of taxa from the northern climate space, with the exception of *Tsuga*.

@@

Some of these apparent shifts in climate niches may be due to differences in sampling design rather than to actual shifts in species distributions.  In particular, the FIA dataset may be underrepresenting tree distributions in unforested or semi-forested regions.  If so, estimates of forest loss (`r fig_nums('gain_loss', display='cite')`) and niche shifts (Figures 4 & 5) may be overestimated. Regardless of cause, the differences between the PLS and FIA tree distributions are important for distributional modeling because FIA data provides the most detailed and most widely used source of tree distributional data over the entire contiguous United States [@wang2016landscape;@nieto2015close;@bell2014early].  Hence, any incompleteness in the representation of realized climate niches based on FIA data that are affecting this study should also affect ecological assessments that are based on FIA data.

>>>>

Contemporary forest inventory data are widely used to understand environmental controls on tree species distributions and to construct models to project forest responses to climate change, but the stability and representativeness of contemporary tree-climate distributions is poorly understood. Here we show that tree-climate relationships for 15 common tree genera in the north-central US have significantly altered over the last two centuries due to historical land use and climate change. Realized niches have shifted towards higher minimum temperatures, cooler maximum temperatures, and higher rainfall. A new attribution method implicates both historical climate change and land use in these shifts, with the relative importance varying among genera and climate variables. Most climate/land-use interactions are compounding, in which historical land-use reinforces climatic shifts in species distributions toward cooler or wetter distributions. The prevalence of compounding interactions imply that contemporary-based models of species distributions may underestimate species resilience to climate change.

@@

Tree distributions and community composition in the upper Midwestern US have been heavily influenced by land use conversion associated with EuroAmerican Settlement in the 19^th^ and 20^th^ Centuries [@goring2016preset;@schulte2007homogenization;@hanberry2012comparison;@rhemtulla2009historical], outbreaks of exotic pathogens [@barnes1979stem], climate changes accompanying the end of the Little Ice Age [@hotchkiss2007response] and current anthropogenic warming [@wicci2011climate]. Modern forests, as represented by FIA surveys [@woudenberg2010forest], show greater homogeneity, weaker ecotones, and significantly different species associations than forests surveyed as part of the Public Land Survey System (PLS) prior to widespread land-use conversion in the late 1800s [@goring2016preset;@schulte2007homogenization]. Rhemtulla et al. [-@rhemtulla2009historical] estimate that *Pinus strobus* and *Tsuga canadensis* now occupy only 4% of their original coverage in Wisconsin. Wildland ecosystems have been converted to agricultural and urban land use [@rhemtulla2009historical], particularly in historical prairie and savanna ecosystems (`r fig_nums("land_use_clim", display = "cite")`f), where almost 50% of southern deciduous savanna was converted to cropland [@rhemtulla2007regional].  In northern forested regions, ranges of early successional species have expanded and abundances of late-successional species are reduced due to widespread logging in the 19^th^ and 20^th^ century, and ongoing land management [@schulte2007homogenization;@goring2016preset].  Additionally, the formerly dominant *Ulmus americana* has experienced rangewide population collapses caused by the introduction of exotic pathogens [@barnes1979stem], as did *Castanea dentata*, with other tree species currently threatened (*Tsuga canadensis*, *Fraxinus* spp.).

Here we demonstrate significant shifts in the climatic distributions of tree genera in the upper Midwest over the last 100 years as a result of historical climate change and land use, and we diagnose the relative influence of historical climate and land use changes on these shifts.  We use gridded historical and contemporary climate observations (PRISM-LT) and estimates of pre-settlement and modern forest vegetation [@goring2016preset] to map the geographic and climatic distributions for 15 major tree taxa in the upper Midwestern United States.  We construct four observed and hypothetical sets of tree-climate relationships, using a 2x2 factorial of pre-settlement and modern vegetation and climates.  We develop a new method, based on Hellinger distances among tree-climate distributions, to quantify the total change and attribute the relative importance of historical vegetation and climate change. We assess whether climate and vegetation changes are compounding, counteracting or confounding based on the magnitude and direction of species-climate shifts. We use this framework to discuss the potential impacts of historical land-use and climate change on inferences about species resilience to climate change, as predicted by species-distribution models based soley on modern distributional data.

@@

Estimates of settlement-era forest composition and tree distribution are based on survey data from the PLS [@white1983history;@schulte2001original;@goring2016preset], aggregated to a 64 km^2^grid across the upper Midwestern United States [@goring2016preset]. We briefly review the methods here. PLS data originally were collected on a 1.61 x 1.61 km (1 x 1 mi) grid across the Upper Midwest from 1830 to   1910 [@schulte2001original;@almendinger1996minnesota;@white1983history]. At each survey point surveyors noted the closest two to four trees and recorded their common name, distances from survey points, diameters at breast height, and azimuths to each tree. Goring et al. [-@goring2016preset] aggregated this data to an 8x8 km grid, resulting in a gridded data product with an average of 61 survey points per cell, or approximately 120 trees per cell. This process included data cleaning, taxonomic standardization, and the application of spatially varying correction factors [@goring2016preset]. Using this data, we produce presence-absence distributions for each taxon, across the upper Midwest for 15 common tree genera. A few common names, such as "Ironwood" can not be clearly resolved to a single tree genera and so we combine *Ostrya* and *Carpinus*, and *Thuja* and *Juniperus*. We define 'Presence' as at least one recorded tree in any 8x8 km grid cell and use the term 'range' as the extent in climate space of all cells with recorded presence for a specific taxon.

@@

Comparisons between historical and modern vegetation must address differences in sampling design between the PLS and FIA datasets [@goring2016preset;@kronenfeld2014validating;@kronenfeld2010influence]. Here, the primary solution for maximizing comparability is via aggregation, both spatially, by aggregating records from both datasets to a common scale (8x8 km), and taxonomically by aggregating taxa to the level of genera. Nonetheless, differences persist.  The FIA contains fewer plots per 8 km cell, with more trees per plot than the PLS data and FIA observations are collected only from forested or partially forested landscapes [@woudenberg2010forest], causing many tree taxa to have smaller spatial extents in the FIA than in the PLS datasets. Scattered trees may be present in regions without permanent or temporary FIA plots, complicating the ability to assign 'absence' within the FIA. However, the practice of using FIA data as a surrogate for tree distribution data is widespread [@iverson2013tree;@iverson1998predicting], so these complications are general. In prior comparative analyses of the PLS and FIA data, Goring et al. [-@goring2016preset] tested the effect of differential sampling design on the mapped patterns of novel and disappearing forests, and found little effect. FIA sampling represents a system with the potential for high heterogeneity within cells because of intensive sampling at the plot level, and, with one to few plots within an 8 x 8km grid cell, high between cell heterogeneity [@goring2016preset]. PLS sampling is likely to have low heterogeneity within and between cells because of the uniform spatial distribution, and the high number of points within cells [@goring2016preset].

Detectability at the plot level for FIA data is likely to be higher than for the PLS, however, given the variable radius of PLS plots, detectability of trees at forest margins is likely to be higher in the PLS, and PLS sample plots are likely to sample higher landscape heterogeneity within a given cell because of their uniform distribution across the landscape.  In mapping tree distributions in the Upper Midwest we consider three classes: "gain", where a tree species was present within a grid cell in the FIA, but absent in the PLS; "loss", where a tree was present in a grid cell during the PLS but absent in the FIA; and "continuous presence", where a tree was present within a grid cell in both FIA and PLS eras.  For all species, the range within the Upper Midwest represents an incomplete sample of the species range, however, the prairie-forest boundary along the western margin represents a southern and western range limit for many species, and represents a climatic space that is warmer and drier than elsewhere in the region.

@@

We use monthly PRISM LT data for 1895 to 2014 CE [@prism2004] to generate historical and modern climate data. The PRISM climate variables are available at 800m resolution and have been interpolated from station locations using elevation, aspect and other data.  We resample estimates for mean daily July temperature ($T_{max}$), mean daily January temperature ($T_{min}$), annual temperature range ($T_{diff}$: $T_{max}$ - $T_{min}$), and annually summed daily precipitation ($P_{ann}$) to the 8 x 8km grid used for PLS and FIA data. These climate variables were chosen because of their relative importance as demonstrated in prior research [@iverson1998predicting] and their use in past climate reconstruction [@maiorano2013building].  Recent statistical advances have opened the possibility of obtaining temperature and precipitation estimates from historical observations as early as the 1830s [@tiptoninreview], however the behaviour of uncertainties for derived variables such as potential evapo-transpiration, which ranked highly in previous work [@iverson1998predicting] are not currently known.

Two periods were selected for calculation of climate normals: 1895 to 1919 for the historical period and 1990 - 2014 for the modern period.  There is an unavoidable temporal mismatch between the historical climate data and the PLS survey period, on the order of 50 to 70 years; few meteorological records exist in this region prior to 1895 CE, and no gridded dataset exists at high spatio-temporal resolution for this time period for all climate variables.  The effects of temporal mismatch should be minor given that Northern Hemisphere temperature trends were generally small during the 19^th^ century [@mann2009global].  To check whether an 1895 to 1919 CE climate normal is a reasonable surrogate for pre-settlement climates we compare the instrumental record against dendroclimatic reconstructions of the Palmer Drought Severity Index [@heim2002review] (`r fig_nums("land_use_clim", display = "cite")`).  Baker et al. [-@baker1985minnesota] show long term increases in mean annual temperature at Ft. Snelling and other spatially proximate records, while corrected estimates from Fort Leavenworth in Kansas [@burnette2010daily] show increasing temperatures in all seasons, with faster increases during winter (0.10^o^C decade^-1^) than during summer (0.07^o^C decade^-1^).  This implies that historical estimates may be biased by +0.35^o^C for $t_{max}$, by +0.5^o^C for $t_{min}$ and +0.15^o^C for $t_{diff}$.  The Modern climate data uses a climate normal that encompases the period of observation  (2000 to 2015) for the FIA data used in this study.

@@

Distributions of tree genera within climate space are described as univariate probability density functions (PDFs) of tree species presence ($V$) along a climate ($C$) gradient, estimated using an unweighted Gaussian kernel density estimator, using R's `density` function.  Differences among PDFs are quantified using Hellinger distances [@matthes2016benchmarking] for the univariate PDFs representing the four combined states of vegetation and climate ($V_{H}C_{H}$, $V_{H}C_{M}$, $V_{M}C_{H}$, $V_{M}C_{M}$, where $M$ and $H$ represent modern and historical conditions). Hellinger distance for two discrete probability distributions is defined as:

$H(P, Q) = \frac{1}{\sqrt{2}} \; \sqrt{\sum_{i=1}^k (\sqrt{p_i} - \sqrt{q_i})^2}$

where $P$ and $Q$ are the distributions, with a common discrete index $i$.

The total shift ($d_{tot}$) is calculated as the Hellinger distance between $V_{M}C_{M}$ and $V_{H}C_{H}$ (`r fig_nums("fourPanel", display = "cite")`a: top left to bottom right). These conditions are observed or recorded in data.  The conditions $V_{M}C_{H}$ and $V_{H}C_{M}$ are synthetic, generated by superimposing historical climate patterns on modern vegetation distributions and by superimposing modern climate on historical distributions.  The effect of climate change on shifting species PDFs is described as $d_{c}$, the difference between PDFs using the early climate normals ($C_{H}$) and modern climate normals ($C_{M}$) overlaid on PLS vegetation (e.g., $V_{H}C_{H}$ versus $V_{H}C_{M}$; `r fig_nums("fourPanel", display = "cite")`a; left column). We interpret PDF changes estimated from vegetation change ($V_{H}$ to {$V_{M}$) superimposed on  historical climate (*e.g.*, $V_{H}C_{H}$ versus $V_{M}C_{H}$; modeled by a shift along the top row of Figure 2a) to indicate change in vegetation attributed to land use change, or, more precicely, the land use change associated with EuroAmerican settlement.  We describe the Hellinger distance between $V_{H}C_{H}$ and $V_{M}C_{H}$ as $d_{v}$.

@@

## Analysis

All analysis uses R [@Team2014], RMarkdown [@rmarkdown2016] and R packages `colorspace`, `captioner`, `raster`, `ggplot2`, `mgcv`, `reshape2`, `plyr`, `gridExtra`, `rgdal`, `RColorBrewer`, `analogue`, and `maptools`, `dplyr` and `purrr`. All code is hosted on GitHub (http://github.com/PalEON-Project/Composition_Climate). Maps use the public domain Natural Earth Data (http://naturalearthdata.com) products for political and landscape features.

@@

`r fig_nums("boxplots", "Changes in the realized climate distributions for tree genera in the upper Midwestern United States. In the box-and-whisker plots, the central line indicates the median, box margins indicate 25^th^ and 75^th^ percentiles, whisker tips indicate 5^th^ and 95^th^ percentiles, and dots indicate outliers beyond the 95^th^ percentile. These shifts in realized climate distributions are produced by the joint effects of historical climate change and land-use change.")`

@@

`r fig_nums("hellPlot", "The total shift in climatic distribution over the last 100 to 150 years for each tree genus ($d_{tot}$) plotted against the amount of shift caused by historical climate change vs. changes in tree distributions ($d_{c} - d_{v}$), the latter mainly due to historical land-use.  Results are plotted separately for each of the four climate variables considered here. Shifts in distributions are calculated as the Hellinger distance between tree-climate distributions for different combinations of pre-settlement and modern vegetation and climate (Fig. 2).  For $d_{c}-d_{v}$ (Y-axis), negative values indicate a stronger influence of historical land-use on shifting climate distributions, while positive values indicate a stronger influence of historical climate change on shifting distributions.")`

@@

`r tab_nums("compTable", "Compounding ($\\Box$), confounding (-) and counteracting ($\\bigcirc$) relationships between land-use and climate change.  When the land-use/climate interaction was regressed against spatial variables (latitude, longitude) no significant relationship was found.")`

@@

pander::pandoc.table(conf_table, justify = "left", style = "rmarkdown")

@@

In this study, shifts in realized climate niches and the underlying interactions between climate change and land use emerge from three basic processes.  First, climates have changed over the 20^th^ and early 21^st^ centuries, with rising winter minimum temperatures, declining summer temperatures, and higher precipitation.  These climate changes are produced by a combination of natural variability and anthropogenic change; the latter signal is strongest in the second half of the 20^th^ century and early 21^st^ century [@estrada2013statistically]. Second, tree species distributions have not yet fully adjusted to these climate changes, causing tree distributions to be in partial disequilibrium with climate [@svenning2013disequilibrium;@williamsinpress]. Third, anthropogenic land use has selectively eliminated portions of geographic and environmental space from species' potential ranges (Figure 1f). Agricultural conversion has largely eliminated open forests in warm and dry regions [@rhemtulla2007], while summer temperatures have declined and precipitation increased, resulting in a compounding effect; species occupy cooler habitats with higher precipitation than during historical times. Intensification of land-use in the northern prairie-forest margin shifts species further than expected along the precipitation gradient. While fire and logging have played a role in changing distribution patterns in the north and northeast, they have not resulted in the exclusion or extirpation of taxa from the northern climate space, with the exception of *Tsuga*.

@@

Some of these apparent shifts in climate niches may be due to differences in sampling design rather than to actual shifts in species distributions. In particular, the FIA dataset may be underrepresenting tree distributions in unforested or semi-forested regions. If so, estimates of forest loss (`r fig_nums('gain_loss', display='cite')`) and niche shifts (Figures 4 & 5) may be overestimated, in particular along the warmest and driest region of the tree distributions in the Midwest. Regardless of cause, the differences between the PLS and FIA tree distributions are important for distributional modeling because FIA data provides the most detailed and most widely used source of tree distributional data over the entire contiguous United States [@wang2016landscape;@nieto2015close;@bell2014early]. Hence, any incompleteness in the representation of realized climate niches based on FIA data that are affecting this study should also affect ecological assessments that are based on FIA data. As improvements in remote sensing and statistical reconstructions of past forested landscapes improve it may become possible to fully explore this issue, but at present the extent of the problem is unclear, although this contribution points to possible effects of incomplete sampling.

Artifacts of the choice of climate normal (1995 - 2014) for the modern data mean an exaggerated effect for $t_{max}$, however, supplemental analysis shows that the patterns of change for the other variables, $t_{min}$, $t_{diff}$ and $p_{ann}$ remain robust. Analysis using pollen-based climate reconstruction shows significant shifts in pollen-climate relationships [@jacques2008pre] following Euro-American settlement in the region providing further support that modern vegetaton-climate relationships have changed since settlement.
